---
title: "02.preprocessing_missing_imputation"
output:
  pdf_document: default
  html_document: default
---

This part audits missing data patterns and prepares imputation outputs. 

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)

# Load cleaned objects produced by 01_setup.Rmd
baseline <- readRDS("data/processed/baseline_clean.rds")
long <- readRDS("data/processed/long_clean.rds")

# Output folders
dir.create("outputs", showWarnings = FALSE, recursive = TRUE)
dir.create("outputs/missing", showWarnings = FALSE, recursive = TRUE)

# Quick sanity checks
cat("Loaded cleaned datasets.\n")
cat("baseline dim:", dim(baseline), "\n")
cat("long dim:", dim(long), "\n")
cat("Unique subjects baseline:", n_distinct(baseline$SUBJECT_ID), "\n")
cat("Unique subjects long:", n_distinct(long$SUBJECT_ID), "\n")
```

Missingness audit: we compute the missing rate for each variable in both the baseline and the longitudinal datasets and export the results as CSV tables for documentation and reporting purposes.
```{r missingness-audit}
# Baseline missing rate per variable
miss_base <- baseline %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_rate") %>%
  arrange(desc(missing_rate))

cat("\nTop 10 baseline missing rates:\n")
print(head(miss_base, 10))

readr::write_csv(miss_base, "outputs/missing/missing_baseline.csv")

# Long missing rate per variable
miss_long <- long %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_rate") %>%
  arrange(desc(missing_rate))

cat("\nTop 10 longitudinal missing rates:\n")
print(head(miss_long, 10))

readr::write_csv(miss_long, "outputs/missing/missing_longitudinal.csv")
```
Missingness by timepoint (CYCLE x MONTH) for key longitudinal outcomes (MMDs, HIT-6, HADS-A/D, INT and MIDAS when available), in order to assess whether missing data patterns varied across follow-up visits.
```{r}
# Define key outcomes (use only those that exist)
key_vars <- c("MMDs", "HIT6", "HADSA", "HADSD", "INT", "MIDAS")
key_vars <- intersect(key_vars, names(long))

# Compute missingness by cycle and month
miss_by_time <- long %>%
  group_by(CYCLE, MONTH) %>%
  summarise(
    across(all_of(key_vars), ~ mean(is.na(.)), .names = "miss_{.col}"),
            .groups = "drop"
    ) %>%
  arrange(CYCLE, MONTH)

cat("\nMissingness by cycle/month (first rows):\n")
print(head(miss_by_time, 10))

# Save for reporting
readr::write_csv(miss_by_time, "outputs/missing/missing_by_cycle_month.csv")

```
MICE diagnostics: md.pattern on a baseline subset
```{r mdpattern-run, fig.width=10, fig.height=6, message=FALSE, warning=FALSE}
if (!requireNamespace("mice", quietly = TRUE)) install.packages("mice")
library(mice)

# Select variables (only those available)
pattern_vars <- intersect(
  c("AGE_OF_ONSET", "BMI", "HEIGTH", "FAMILIARITY", "Aura", "Psycopathological"),
  names(baseline)
)

cat("\nVariables used for md.pattern:\n")
print(pattern_vars)

if (length(pattern_vars) >= 2) {

  tmp_plot <- baseline %>% select(all_of(pattern_vars))

  # Create short labels for plotting (only rename columns that exist)
  short_names <- c(
    "AGE_OF_ONSET" = "onset",
    "BMI" = "bmi",
    "HEIGTH" = "height",
    "FAMILIARITY" = "fam",
    "Aura" = "aura",
    "Psycopathological" = "psycho"
  )

  # Apply renaming safely
  tmp_plot <- tmp_plot %>%
    rename_with(~ unname(short_names[.x]),
                .cols = intersect(names(tmp_plot), names(short_names)))

  # Make text smaller and increase top margin
  op <- par(cex = 0.9, mar = c(5, 4, 6, 2) + 0.1)

  md.pattern(tmp_plot)

  par(op)

} else {
  cat("\nNot enough variables found for md.pattern.\n")
  cat("Available baseline columns are:\n")
  print(names(baseline))
}

```


Export md.pattern table: this step saves the md.pattern output as a table. It is useful for reporting missingness patterns without relying on the plot. 
```{r}
if (length(pattern_vars) >= 2) {

  # Recreate the same temporary dataset used for plotting
  tmp <- baseline %>% select(all_of(pattern_vars))

  # Short variable names (same as Step 4)
  short_names <- c(
    "AGE_OF_ONSET" = "onset",
    "BMI" = "bmi",
    "HEIGTH" = "height",
    "FAMILIARITY" = "fam",
    "Aura" = "aura",
    "Psycopathological" = "psycho"
  )

  # Apply renaming safely (only existing columns)
  tmp <- tmp %>%
    rename_with(
      ~ unname(short_names[.x]),
      .cols = intersect(names(tmp), names(short_names))
    )

  # Compute md.pattern without plotting
  md_tab <- md.pattern(tmp, plot = FALSE)

  # Save table to CSV for the report
  write.csv(md_tab,
            "outputs/missing/mdpattern_baseline_subset.csv",
            row.names = TRUE)

  cat("\nmd.pattern table saved in outputs/missing/mdpattern_baseline_subset.csv\n")

} else {
  cat("\nStep 5 skipped: not enough variables for md.pattern export.\n")
}
```
Baseline imputation (MICE)
After inspecting missingness patterns in a subset of baseline variables (md. pattern), we run a multivariate imputation by chained equations (MICE) model on the selected baseline covariates.

The goal is to obtain one completed baseline dataset ('baseline_imp') that we will later merge with the longitudinal data for the main analyses.

```{r mice-baseline-main, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
library(mice)

mice_vars <- intersect(
  c("SUBJECT_ID", "AGE", "BMI", "WEIGTH", "HEIGTH", "AGE_OF_ONSET", "FAMILIARITY", "Aura", "Psycopathological"),
  names(baseline)
)

base_mice <- baseline %>% dplyr::select(all_of(mice_vars))

# Convert categorical vars to factors (only if present)
cat_vars <- intersect(c("FAMILIARITY", "Aura", "Psycopathological"), names(base_mice))
base_mice <- base_mice %>% mutate(across(all_of(cat_vars), as.factor))

# Do not impute ID
meth <- make.method(base_mice)
meth["SUBJECT_ID"] <- ""

pred <- make.predictorMatrix(base_mice)
pred[, "SUBJECT_ID"] <- 0
pred["SUBJECT_ID", ] <- 0

set.seed(123)
imp <- mice(base_mice, m = 5, maxit = 20, method = meth, predictorMatrix = pred)

plot(imp)  # trace plots

if ("BMI" %in% names(base_mice)) densityplot(imp, ~ BMI)
if ("AGE_OF_ONSET" %in% names(base_mice)) densityplot(imp, ~ AGE_OF_ONSET)

baseline_imp <- complete(imp, 1)

dir.create("data/processed", showWarnings = FALSE, recursive = TRUE)
saveRDS(imp, "data/processed/baseline_mice_object.rds")
saveRDS(baseline_imp, "data/processed/baseline_imputed.rds")

```
MICE ran for 20 iterations with 5 imputed datasets. We inspected convergence using trace plots and checked the plausibility of imputations using density plots for continuous variables.

We then extracted one completed baseline dataset ('baseline_imp = complete(imp, 1)') for the main analysis workflow and saved both the MICE object and the imputed baseline dataset as '.rds' files:
- 'data/processed/baseline_mice_object.rds'
- 'data/processed/baseline_imputed.rds'

Diagnostic figures were generated in the rendered output for reporting. 

===============================================================================
**Longitudinal Imputation (main and sensitivity analyses):**
In this step, we address missing values in the longitudinal outcomes.
Given the repeated-measures structure of the data and the presence of distinct treatment cycles separated by a washout (suspension) period, imputation is performed **within each patient and within each treatment cycle**, without borrowing information across cycles.

We define two longitudinal datasets:
- a **main dataset**, using linear interpolation between observed values;
- a **sensitivity dataset**, using a last observation carried forward (LOCF) approach.

In both cases, missing values occurring **before the first observed measurement within a cycle** are not imputed, in order to avoid introducing artificial baseline values.

```{r longitudinal-imputation, message=FALSE, warning=FALSE}
if (!requireNamespace("zoo", quietly = TRUE)) install.packages("zoo")
library(zoo)

# Key longitudinal variables to impute (keep only those that exist)
vars_long_imp <- c("MMDs", "HIT6", "HADSA", "HADSD", "INT", "MIDAS")
vars_long_imp <- intersect(vars_long_imp, names(long))

cat("\nLongitudinal variables considered for imputation:\n")
print(vars_long_imp)

# Ensure correct ordering (within subject and cycle)
long_sorted <- long %>%
  arrange(SUBJECT_ID, CYCLE, MONTH)

# -------------------------
# MAIN: linear interpolation within SUBJECT_ID x CYCLE
# - na.approx does NOT extrapolate when na.rm=FALSE:
#   leading NAs and trailing NAs remain NA
# -------------------------
long_main <- long_sorted %>%
  group_by(SUBJECT_ID, CYCLE) %>%
  arrange(MONTH, .by_group = TRUE) %>%
  mutate(
    across(
      all_of(vars_long_imp),
      ~ zoo::na.approx(., x = MONTH, na.rm = FALSE),
      .names = "{.col}"
    )
  ) %>%
  ungroup()

# -------------------------
# SENSITIVITY: LOCF within SUBJECT_ID x CYCLE
# - carries last observed value forward
# - leading NAs remain NA when na.rm=FALSE
# -------------------------
long_sens <- long_sorted %>%
  group_by(SUBJECT_ID, CYCLE) %>%
  arrange(MONTH, .by_group = TRUE) %>%
  mutate(
    across(
      all_of(vars_long_imp),
      ~ zoo::na.locf(., na.rm = FALSE),
      .names = "{.col}"
    )
  ) %>%
  ungroup()

# --- Sanity check: do not impute before the first observation within each cycle ---
check_leading <- long_sorted %>%
  group_by(SUBJECT_ID, CYCLE) %>%
  arrange(MONTH, .by_group = TRUE) %>%
  summarise(
    lead_na_MMDs_orig = is.na(first(MMDs)),
    lead_na_MMDs_main = is.na(first(long_main$MMDs[match(paste(SUBJECT_ID, CYCLE, first(MONTH)),
                                                        paste(long_main$SUBJECT_ID, long_main$CYCLE, long_main$MONTH))])),
    lead_na_MMDs_sens = is.na(first(long_sens$MMDs[match(paste(SUBJECT_ID, CYCLE, first(MONTH)),
                                                        paste(long_sens$SUBJECT_ID, long_sens$CYCLE, long_sens$MONTH))])),
    .groups = "drop"
  )

print(head(check_leading, 10))

write.csv(check_leading,
          "outputs/missing/leading_na_check_MMDs.csv",
          row.names = FALSE)


# Any violations? (leading NA in original should remain NA after imputation)
viol_main <- sum(check_leading$lead_na_MMDs_orig & !check_leading$lead_na_MMDs_main, na.rm = TRUE)
viol_sens <- sum(check_leading$lead_na_MMDs_orig & !check_leading$lead_na_MMDs_sens, na.rm = TRUE)

cat("\nLeading-NA violations (MMDs):\n")
cat("main (interp):", viol_main, "\n")
cat("sens (LOCF):", viol_sens, "\n")

# Save for downstream steps
dir.create("data/processed", showWarnings = FALSE, recursive = TRUE)
saveRDS(long_main, "data/processed/long_imputed_main_interp.rds")
saveRDS(long_sens, "data/processed/long_imputed_sens_locf.rds")

cat("\nSaved:\n- data/processed/long_imputed_main_interp.rds\n- data/processed/long_imputed_sens_locf.rds\n")
```
After imputation, we performed a sanity check to verify that no values were imputed before the first observed measurement within each treatment cycle.

Specifically, we compared the presence of leading missing values in the original data with the imputed datasets for the main (interpolation) and sensitivity (LOCF) approaches.

No leading-NA violations were detected for primary outcomes (MMDs), confirming that the imputation respected the cycle structure and did not borrow information from previous cycles or introduce artificial baseline values.

===============================================================================
**Final merge of baseline and longitudinal data**

In this step, we merge the imputed baseline dataset with the longitudinal datasets obtained after missing-data handling.

this results into two final analysis-ready datasets:
- a **main dataset**, based on linear interpolation of longitudinal outcomes;
- a **sensitivity dataset**, based on LOCF imputation.

Each dataset contains one row per subject x cycle x month, enriched with subject-level baseline covariates and will be used in downstream mixed-effects and  mismatch analyses.

```{r merge-final-datasets, message=FALSE, warning=FALSE}
# Load imputed datasets
baseline_imp <- readRDS("data/processed/baseline_imputed.rds")
long_main <- readRDS("data/processed/long_imputed_main_interp.rds")
long_sens <- readRDS("data/processed/long_imputed_sens_locf.rds")

# Merge baseline covariates into longitudinal data
data_main <- long_main %>%
  left_join(baseline_imp, by = "SUBJECT_ID")

data_sens <- long_sens %>%
  left_join(baseline_imp, by = "SUBJECT_ID")

# Sanity checks
cat("Rows (main):", nrow(data_main), "\n")
cat("Rows (sens):", nrow(data_sens), "\n")

cat("Unique subjects (main):", n_distinct(data_main$SUBJECT_ID), "\n")
cat("Unique subjects (sens):", n_distinct(data_sens$SUBJECT_ID), "\n")

cat("Missing baseline after merge (main):",
    sum(is.na(data_main$AGE)), "\n")

# Check for duplicated visit identifiers
dup_main <- data_main %>%
  count(SUBJECT_ID, CYCLE, MONTH) %>%
  filter(n > 1)

dup_sens <- data_sens %>%
  count(SUBJECT_ID, CYCLE, MONTH) %>%
  filter(n > 1)

cat("Duplicate visits (main):", nrow(dup_main), "\n")
cat("Duplicate visits (sens):", nrow(dup_sens), "\n")

# Save final datasets
saveRDS(data_main, "data/processed/data_main_analysis.rds")
saveRDS(data_sens, "data/processed/data_sens_analysis.rds")

cat("\nSaved final datasets:\n")
cat("- data/processed/data_main_analysis.rds\n")
cat("- data/processed/data_sens_analysis.rds\n")
```



The merge did not introduce duplicated visits or inconsistencies in the longotudinal structure.
The resulting datasets are now fully analysis-ready and will be sued for mixed-effects modeling and mismatch analyses addressing Task 1 and Task 2.