---
title: "2_task"
output: html_document
---

Task 2: Factors Associated with Objective-Subjective Discordance (CYCLE 1)

The aim of Task 2 is to investigate why some patients show an objective improvement under anti_CGRP treatment (reduction in Monthly Migraine Days, MMDs) without experiecing a clinically meaningful improvement in patient-reported impact (HIT-6).

In Task 1 we established that this objective-subjective mismatch exists and is robust across alternative missing-data strategies. Task 2 moves from description to explanation by assessing whether baseline clinical and psychosocial characteristics are associated with the discordant profile.

A crucial clinical distinction is that there are two opposite discordance patterns:
- Unhappy responder (primary focus): objective responder but subjective non-responder (OBJ+ / Subj-).
- Happy Non-Responder (secondary/opposite pattern): objective non-responder but subjecctive responder (Obj- / Subj+).

Because these profile reflect different mechanisms, the main analysis in Task 2 will focus on the Unhappy Responder outcome, rather than pooling all mismathes into a signle indicator.

**Setup**

```{r setup, message=FALSE, warning=FALSE}
# ------------------------------------------------------------
# Setup (Task 2)
# Goal: ensure we are at the PROJECT ROOT so relative paths work.
# ------------------------------------------------------------

if (!requireNamespace("rstudioapi", quietly = TRUE)) install.packages("rstudioapi")
library(tidyverse)

# Project root (works in RStudio projects)
project_root <- getwd()
if (rstudioapi::isAvailable()) {
  prj <- rstudioapi::getActiveProject()
  if (!is.null(prj) && nzchar(prj)) project_root <- prj
}

cat("project_root:", project_root, "\n")
cat("getwd():", getwd(), "\n")

# Output folders for Task 2 (always under PROJECT ROOT)
dir.create(file.path(project_root, "outputs", "task2"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(project_root, "outputs", "task2", "figures"), recursive = TRUE, showWarnings = FALSE)


# Quick path check (this prevents silent path bugs)
stopifnot(dir.exists("outputs"))
stopifnot(dir.exists("outputs/task1"))

# ------------------------------------------------------------
# Load patient-level dataset created in Task 1 (MAIN analysis)
# ------------------------------------------------------------

file_c1_main <- file.path("outputs", "task1", "cycle1_responder_discordance_main.csv")
cat("Trying to read:", file_c1_main, "\n")
stopifnot(file.exists(file_c1_main))

c1_main <- readr::read_csv(file_c1_main, show_col_types = FALSE)

# Quick sanity checks
cat("Rows (patients):", nrow(c1_main), "\n")
cat("Unique subjects:", dplyr::n_distinct(c1_main$SUBJECT_ID), "\n")

# Ensure expected columns exist (adapt if your names differ)
stopifnot(all(c("SUBJECT_ID", "obj_resp", "subj_resp", "discordance") %in% names(c1_main)))

# ------------------------------------------------------------
# End of setup: now paths + base dataset are ready.
# ------------------------------------------------------------

```
This setup chunk initializes the environment, creates output folders for Task 2, and loads the patient-level Cycle 1 dataset generated in Task 1. The dataset already contains objective and subjective responder indicators (obj_resp, subj_resp), which are required to define the discordance outcomes. 

=================================================================
**Define discordance outcocmes + counts **
We now create two distinct binary outcomes to avoid mixing opposite discordance phenomena. The main outcome for the logistic model will be the Unhappy Responder profile (Obj+ / Subj-).

```{r task2-step20-outcome, message=FALSE, warning=FALSE}
# ------------------------------------------------------------
# Define outcome profiles
# ------------------------------------------------------------

c1_main <- c1_main %>%
  mutate(
    unhappy_responder = as.integer(obj_resp == 1L & subj_resp == 0L),
    happy_nonresponder = as.integer(obj_resp == 0L & subj_resp == 1L)
  )

# Counts
tab_unhappy <- table(c1_main$unhappy_responder, useNA = "ifany")
tab_happy <- table(c1_main$happy_nonresponder, useNA = "ifany")

cat("Unhappy responders (1=yes):\n"); print(tab_unhappy)
cat("\nHappy non-responders (1=yes):\n"); print(tab_happy)

# Summary table
outcome_counts <- tibble(
outcome = c("unhappy_responder", "happy_nonresponder"),
n_yes = c(sum(c1_main$unhappy_responder == 1, na.rm = TRUE),
sum(c1_main$happy_nonresponder == 1, na.rm = TRUE)),
n_total = c(nrow(c1_main), nrow(c1_main)),
pct_yes = 100 * n_yes / n_total
)
outcome_counts
readr::write_csv(outcome_counts, "outputs/task2/outcome_counts.csv")
```
We created two separate discordance outcomes. The unhappy-responder variable identifies patients who meet the objective responder criterion, but do not achieve a clinically meaningful HIT-6 improvement (the clinically central mismatch).
The happy_nonresponder variable captures the opposite pattern. We also report counts and percentage to assess how many events are available for subsequent logistic modeling (important avoiding overfitting)

```{r}
# ------------------------------------------------------------
# Load Task 1 patient-level dataset
# ------------------------------------------------------------

c1_task1 <- readr::read_csv(
  "outputs/task1/cycle1_responder_discordance_main.csv",
  show_col_types = FALSE
)

# Sanity checks
cat("Patients:", nrow(c1_task1), "\n")
cat("Unique SUBJECT_ID:", dplyr::n_distinct(c1_task1$SUBJECT_ID), "\n")

# Check required columns
stopifnot(all(c("SUBJECT_ID", "obj_resp", "subj_resp") %in% names(c1_task1)))
```
We start from the patient-level dataset generated in Task 1, which contains one row per subject with objective and subjective responder indicators at the end of Cycle 1.
This dataset represents the analytical backbone for Task2 and ensures that all subsequent analyses are conducted at the patient level.
