---
title: "2_task"
output: html_document
---

Task 2: Factors Associated with Objective-Subjective Discordance (CYCLE 1)

The aim of Task 2 is to investigate why some patients show an objective improvement under anti_CGRP treatment (reduction in Monthly Migraine Days, MMDs) without experiecing a clinically meaningful improvement in patient-reported impact (HIT-6).

In Task 1 we established that this objective-subjective mismatch exists and is robust across alternative missing-data strategies. Task 2 moves from description to explanation by assessing whether baseline clinical and psychosocial characteristics are associated with the discordant profile.

A crucial clinical distinction is that there are two opposite discordance patterns:
- Unhappy responder (primary focus): objective responder but subjective non-responder (OBJ+ / Subj-).
- Happy Non-Responder (secondary/opposite pattern): objective non-responder but subjecctive responder (Obj- / Subj+).

Because these profile reflect different mechanisms, the main analysis in Task 2 will focus on the Unhappy Responder outcome, rather than pooling all mismathes into a signle indicator.

**Setup**

```{r setup, message=FALSE, warning=FALSE}
# ------------------------------------------------------------
# Setup (Task 2)
# Goal: ensure we are at the PROJECT ROOT so relative paths work.
# ------------------------------------------------------------

if (!requireNamespace("rstudioapi", quietly = TRUE)) install.packages("rstudioapi")
library(tidyverse)

# Project root (works in RStudio projects)
project_root <- getwd()
if (rstudioapi::isAvailable()) {
  prj <- rstudioapi::getActiveProject()
  if (!is.null(prj) && nzchar(prj)) project_root <- prj
}

cat("project_root:", project_root, "\n")
cat("getwd():", getwd(), "\n")

# Output folders for Task 2 (always under PROJECT ROOT)
dir.create(file.path(project_root, "outputs", "task2"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(project_root, "outputs", "task2", "figures"), recursive = TRUE, showWarnings = FALSE)


# Quick path check (this prevents silent path bugs)
stopifnot(dir.exists("outputs"))
stopifnot(dir.exists("outputs/task1"))

# ------------------------------------------------------------
# Load patient-level dataset created in Task 1 (MAIN analysis)
# ------------------------------------------------------------

file_c1_main <- file.path("outputs", "task1", "cycle1_responder_discordance_main.csv")
cat("Trying to read:", file_c1_main, "\n")
stopifnot(file.exists(file_c1_main))

c1_main <- readr::read_csv(file_c1_main, show_col_types = FALSE)

# Quick sanity checks
cat("Rows (patients):", nrow(c1_main), "\n")
cat("Unique subjects:", dplyr::n_distinct(c1_main$SUBJECT_ID), "\n")

# Ensure expected columns exist (adapt if your names differ)
stopifnot(all(c("SUBJECT_ID", "obj_resp", "subj_resp", "discordance") %in% names(c1_main)))

# ------------------------------------------------------------
# End of setup: now paths + base dataset are ready.
# ------------------------------------------------------------

```
This setup chunk initializes the environment, creates output folders for Task 2, and loads the patient-level Cycle 1 dataset generated in Task 1. The dataset already contains objective and subjective responder indicators (obj_resp, subj_resp), which are required to define the discordance outcomes. 

---
**Define discordance outcocmes + counts **
We now create two distinct binary outcomes to avoid mixing opposite discordance phenomena. The main outcome for the logistic model will be the Unhappy Responder profile (Obj+ / Subj-).

```{r task2-step20-outcome, message=FALSE, warning=FALSE}
# ------------------------------------------------------------
# Define outcome profiles
# ------------------------------------------------------------

c1_main <- c1_main %>%
  mutate(
    unhappy_responder = as.integer(obj_resp == 1L & subj_resp == 0L),
    happy_nonresponder = as.integer(obj_resp == 0L & subj_resp == 1L)
  )

# Counts
tab_unhappy <- table(c1_main$unhappy_responder, useNA = "ifany")
tab_happy <- table(c1_main$happy_nonresponder, useNA = "ifany")

cat("Unhappy responders (1=yes):\n"); print(tab_unhappy)
cat("\nHappy non-responders (1=yes):\n"); print(tab_happy)

# Summary table
outcome_counts <- tibble(
outcome = c("unhappy_responder", "happy_nonresponder"),
n_yes = c(sum(c1_main$unhappy_responder == 1, na.rm = TRUE),
sum(c1_main$happy_nonresponder == 1, na.rm = TRUE)),
n_total = c(nrow(c1_main), nrow(c1_main)),
pct_yes = 100 * n_yes / n_total
)
outcome_counts
readr::write_csv(outcome_counts, "outputs/task2/outcome_counts.csv")
```
We created two separate discordance outcomes. The unhappy-responder variable identifies patients who meet the objective responder criterion, but do not achieve a clinically meaningful HIT-6 improvement (the clinically central mismatch).
The happy_nonresponder variable captures the opposite pattern. We also report counts and percentage to assess how many events are available for subsequent logistic modeling (important avoiding overfitting)


We start from the patient-level dataset generated in Task 1, which contains one row per subject with objective and subjective responder indicators at the end of Cycle 1.
This dataset represents the analytical backbone for Task2 and ensures that all subsequent analyses are conducted at the patient level.

---
**Build the Task 2 analytic dataset (patient-level)**
The objective of this step is to construct the final analytic dataset for Task 2, containing one row per patient and combining the discordance outcome defined in Step 2.0 with baseline clinical and demographic predictors.

This dataset will be used as input for the logistic regression models aimed at identifying factors associated with the unhappy responder profile.

**Patient-level dataset from Task 1**
Task 2 starts from the patient-level dataset generated in Task 1, which contains one row per subject and summarizes objective and subjective treatment response at the end of Cycle 1.
This dataset already includes:
- the unique patient identifier (SUBJECT_ID),
- objective responder status based on MMD reduction (obj_resp),
- subjective responder status based on HIT-6 improvement (subj_resp),
- a discordance indicator derived from these two dimensions.

Using this dataset as the starting point ensures that all subsequent analyses are performed at the patient level and are directly linked to the response definitions established in Task 1.

**Addition of baseline predictors**

To investigate which factors are associated with the objective–subjective discordance, baseline clinical and demographic variables are added to the patient-level dataset.

Baseline information is retrieved from the cleaned baseline dataset and merged using SUBJECT_ID. The predictors considered include:
- demographic variables (e.g., age, sex),
- clinical characteristics (diagnosis, comorbidities),
- baseline migraine burden and severity.

A left join is used to preserve the analytic sample defined in Task 1, ensuring that all patients included in the discordance analysis remain in the dataset.


```{r task2-step21-merge-baseline, message=FALSE, warning=FALSE}
# ------------------------------------------------------------
# Load baseline (RDS) + merge into Cycle 1 dataset
# ------------------------------------------------------------

file_baseline <- file.path("data", "processed", "baseline_clean.rds")
cat("Trying to read baseline:", file_baseline, "\n")
stopifnot(file.exists(file_baseline))

baseline <- readRDS(file_baseline)

# safety: 1 row per subject
baseline <- baseline %>% distinct(SUBJECT_ID, .keep_all = TRUE)

# merge (keeps Task 1 analytic sample)
task2_df <- c1_main %>%
  left_join(baseline, by = "SUBJECT_ID")

# sanity checks
cat("Rows task2_df:", nrow(task2_df), "\n")
cat("Unique subjects task2_df:", dplyr::n_distinct(task2_df$SUBJECT_ID), "\n")

# if AGE exists in baseline, this is a good merge check
if ("AGE" %in% names(task2_df)) {
  cat("Missing AGE after merge:", sum(is.na(task2_df$AGE)), "\n")
}

# save
readr::write_csv(task2_df, "outputs/task2/task2_cycle1_analytic_raw.csv")
```

**Four-profile clinical classification**

For descriptive purposes and sanity checks, patients are additionally classified into four mutually exclusive response profiles based on the combination of objective and subjective response:

- Concordant responders (OBJ+ / Subj+),
- Unhappy responders (OBJ+ / Subj−),
- Happy non-responders (OBJ− / Subj+),
- Concordant non-responders (OBJ− / Subj−).

This four-quadrant classification is not used as the primary outcome in the main logistic model, but it provides a clinically intuitive representation of treatment response patterns and helps contextualize the unhappy responder group.


```{r task2-step22-profile4, message=FALSE, warning=FALSE}
# ------------------------------------------------------------
#  4-quadrant profile (OBJ/Subj)
# ------------------------------------------------------------

task2_df <- task2_df %>%
  mutate(
    profile_4 = case_when(
      obj_resp == 1L & subj_resp == 1L ~ "Concordant responder (OBJ+ / Subj+)",
      obj_resp == 1L & subj_resp == 0L ~ "Unhappy responder (OBJ+ / Subj-)",
      obj_resp == 0L & subj_resp == 1L ~ "Happy non-responder (OBJ- / Subj+)",
      obj_resp == 0L & subj_resp == 0L ~ "Concordant non-responder (OBJ- / Subj-)",
      TRUE ~ NA_character_
    ),
    profile_4 = factor(
      profile_4,
      levels = c(
        "Concordant responder (OBJ+ / Subj+)",
        "Unhappy responder (OBJ+ / Subj-)",
        "Happy non-responder (OBJ- / Subj+)",
        "Concordant non-responder (OBJ- / Subj-)"
      )
    )
  )

# Counts + save
profile_counts <- as.data.frame(table(task2_df$profile_4, useNA = "ifany"))
names(profile_counts) <- c("profile_4", "n")
profile_counts$pct <- 100 * profile_counts$n / sum(profile_counts$n)

print(profile_counts)
readr::write_csv(profile_counts, "outputs/task2/profile4_counts.csv")
```